\note{The idea here is to change the geometry, give them some extra practice problems to work on, possibly solutions if they exist, and then restore the geometry. Do this at the end of every chapter.}



\newgeometry{left=1in,right=1in,top=1in,bottom=1in}

\section*{Extra Practice}
\addcontentsline{toc}{section}{Extra Practice}
At the end of each chapter, I'll include some extra practice problems. These problems might come in the form of a reference to problems in Schaum's Outlines  \textit{Differential Equations} by Richard Bronson. Sometimes I'll point you to a freely available open source text. Generally, the problems will come with solutions where you can check your work.  You can also use either Sage or Mathematica to check most solutions.

This chapter consists mostly of a review of concepts from calculus and multivariate calculus. As such, probably the best place to look for review problems is in your old calculus textbook.  At some point, I'll either find, or create, a good collection of practice problems to help you.  As of now, the best I can do is point you to problems in Thomas's Calculus 12 Edition, and Schaum's Outlines.   
\begin{center}
\begin{tabular}{|l|l|l|}
\hline
Concept                      & Source	             & Relevant Problems\\\hline
Differentials                &Thomas's& Section 14.6\\ \hline
Vector Fields                &Thomas's& Section 16.2\\ \hline
Potentials                   &Thomas's& Section 16.3\\ \hline
Separable ODEs               &Schaum's& Chapter 4: 1-8, 23-45\\ \hline
Exact ODEs (find a potential)&Schaum's& Chapter 5: 1-13, 24-40, 56-65\\ \hline
Laplace Transforms           &Schaum's& Chapter 21: 4-7, 10-12, 27-35 (wait)\\\hline
\end{tabular}
\end{center}

\restoregeometry